Demes: a specification for interchanging complex demography

### 2022-04-11 paper.tex
implement suggestions and cleanup

@@

In Demes, demographic models consist of one or more interacting populations,
or ``demes'', understood to be a collection of individuals that can be
conveniently modelled using a defined set of rules and parameters
\citep{gilmour_demes_1939,gilmour_deme_1955}.
To avoid confusion with the name of the specification itself we will
use the term ``population'' in this discussion, with the understanding that the
terms are interchangeable.
A population is defined as some collection of individuals that exists for
some period of time, and has a well-defined size (i.e., number of individuals)
during that time period. Individuals can move between populations
either according to their ancestor-descendant relationships
or through processes involving migrations.
Few other properties of the populations are specified in the model:
we are concerned primarily with defining the populations, their sizes, and the
movement of individuals between those populations.
\aprcomment{previous sentence changed to allow for selfing/cloning}

@@

Within a population, we implicitly assume that allele frequency dynamics can
be described by the Wright-Fisher model.
Briefly, generations are non-overlapping (all parents
reproduce and die simultaneously), and for allele $i$ currently at frequency $p_i$,
its frequency in the next generation (at birth) is expected to be $p_iw_i/\bar{w}$,
where $w_i$ and $\bar{w}$ are the marginal and
mean fitnesses, respectively, properly weighted according to ancestry proportions.
In this framework,
a forward-time simulation of finite populations is equivalent
to multinomial sampling of allele frequencies each
generation (\citet[][pp 29-31]{burger2000-ul}, \citet[][pp 179-181]{crowkimura1970}),
and a backwards-time (coalescent) simulation follows the approximations
described in \citet{tajima1983evolutionary,hudson1983testing}
and \citet[][chapter 3]{wakeley2008-hd}.
Further, this model assumes ``soft'' selection \citep{christiansen1975hard},
meaning that the dynamics of population sizes
changes are \textit{independent} of the details of individual fitnesses.
% In the \texttt{Discussion}, we discuss alternative models compatible
% with this framework and outline the expectations for tools implementing
% them.
\aprcomment{As such, this model excludes scenarios such as ``hard selection,''
in which population sizes are dependent on a population's mean fitness, or
stochastic fluctuations in population size, such as defining ``carrying capacities''.}
Many forwards and backwards time simulators currently implement this model
\citep[e.g.,][\aprcomment{and more\dots?}]{hudson2002generating,gutenkunst2009inferring,excoffier2011fastsimcoal,jouganous2017inferring,haller2019slim,thornton2019-nu}.
Each has their own unique method of specifying the details of selection and demography.
The goal of \Demes\ is to provide a common method for the latter task.

@@

Each population has an assigned selfing rate and cloning rate, where each defines
the probability that offspring are generated from one generation to the next by either
self-fertilisation or cloning of an individual.
More specifically, for a given epoch within a population denote the clonal rate by
$\sigma$ and the selfing rate by $S$.
$S$ and $\sigma$ can take any value between zero and one and
can sum to more than one.
Each generation a proportion of offspring
$\sigma$ are expected to be generated through clonal reproduction,
while $1-\sigma$ are expected to arise through sexual reproduction.
Within the sexually-reproduced offspring,
$S$ are born via self-fertilisation while the rest
have parents drawn at random from the previous generation.
Depending on the simulation software, this random drawing of parent may occur
either with or without replacement. When drawing occurs with replacement, a small
amount of ``residual'' selfing is expected, so that the realized selfing probability
is $(1-\sigma)(S + (1-S)/N)$ instead of $(1-\sigma)S$ (so that even with $\sigma=0$
and $S=0$, selfing may still occur with probability $1/N$).
Simulation software that allow variable rates of selfing are expected to clearly
document the expected behaviour.

By allowing the definition of selfing and cloning probabilities, we allow many 
standard models to be defined.
However, by parameterising selfing and cloning as we have, we assume that these
properties of populations can be specified independently from the genetics.
In other words, mutations that cause selfing probabilities to fluctuate within
an epoch are not considered.
\aprcomment{
    Do we still want to say something along these lines?
    ``However, consuming software is free to ignore these parameters
    and model the genetics within
    the framework of a soft selection model.
    We don't view these as limitations. Rather, this is what most people
    are doing most of the time.
    Just now we can finally try to do this part w/fewer errors.''
}
\krtcomment{
    I think we want a section in the Discussion outlining some obvious things
    that people can do w/demes that are sitting "just outside" the strict
    specification: hard selection, etc..
}
More details of the mathematical properties of selfing and cloning rates
in a coalescent context can be found in \citet{hartfield2016facsexcoal}.

>>>>

\krtcomment{
    I think we want a section in the Discussion outlining some obvious things
    that people can do w/demes that are sitting "just outside" the strict
    specification: hard selection, etc..
}

@@

In Demes, demographic models consist of one or more interacting populations,
or ``demes'', understood to be a collection of individuals that can be
conveniently modelled using a defined set of rules and parameters
\citep{gilmour_demes_1939,gilmour_deme_1955}.
To avoid confusion with the name of the specification itself we will
use the term ``population'' in this discussion, with the understanding that the
terms are interchangeable.
A population is defined as some collection of individuals that exists for
some period of time, and has a well-defined size (i.e., number of individuals)
during that time period. Individuals can move between populations
either according to their ancestor-descendant relationships
or through processes involving migrations.
Few other properties of the populations are specified in the model:
we are concerned primarily with defining the populations, their sizes, and the
movement of individuals between those populations.

@@

Within a population, we assume that allele frequency dynamics can
be described by the Wright-Fisher model.
Briefly, generations are non-overlapping (all parents
reproduce and die simultaneously), and for allele $i$ currently at frequency $p_i$,
its frequency in the next generation (at birth) is expected to be $p_iw_i/\bar{w}$,
where $w_i$ and $\bar{w}$ are the marginal and
mean fitnesses, respectively, properly weighted according to ancestry proportions.
In this framework,
a forward-time simulation of finite populations is equivalent
to multinomial sampling of allele frequencies each
generation (\citet[][pp 29-31]{burger2000-ul}, \citet[][pp 179-181]{crowkimura1970}),
and a backwards-time (coalescent) simulation follows the approximations
described in \citet{tajima1983evolutionary,hudson1983testing}
and \citet[][chapter 3]{wakeley2008-hd}.
Further, this model assumes ``soft'' selection \citep{christiansen1975hard},
meaning that the dynamics of population sizes
changes are \textit{independent} of the details of individual fitnesses.
As such, this model excludes scenarios such as ``hard selection,''
in which population sizes are dependent on a population's mean fitness, or
stochastic fluctuations in population size, such as interpreting
population sizes as ``carrying capacities''.
Many forwards and backwards time simulators currently implement this model
\citep[e.g.,][]{hudson2002generating,gutenkunst2009inferring,excoffier2011fastsimcoal,kelleher2016efficient,jouganous2017inferring,haller2019slim,thornton2019-nu}.
Each has their own unique method of specifying the details of selection and demography.
The goal of \Demes\ is to provide a common method for the latter task.

@@

Each population has an assigned selfing rate and cloning rate, where each defines
the probability that offspring are generated from one generation to the next by either
self-fertilisation or cloning of an individual.
More specifically, for a given epoch within a population denote the clonal rate by
$\sigma$ and the selfing rate by $S$.
$S$ and $\sigma$ can take any value between zero and one and
can sum to more than one.
Each generation a proportion of offspring
$\sigma$ are expected to be generated through clonal reproduction,
while $1-\sigma$ are expected to arise through sexual reproduction.
Within the sexually-reproduced offspring,
$S$ are born via self-fertilisation while the rest
have parents drawn at random from the previous generation.
Depending on the simulation software, this random drawing of parent may occur
either with or without replacement. When drawing occurs with replacement, a small
amount of ``residual'' selfing is expected, so that the realized selfing probability
is $(1-\sigma)(S + (1-S)/N)$ instead of $(1-\sigma)S$ (so that even with $\sigma=0$
and $S=0$, selfing may still occur with probability $1/N$), although
this effect is negligible in large populations \citep{nordborg1997coalescent}
Simulation software that allow variable rates of selfing are expected to clearly
document the expected behaviour.

By allowing the definition of selfing and cloning probabilities, we allow many 
standard models to be defined.
However, by parameterising selfing and cloning as we have, we assume that these
properties of populations can be specified independently from the genetics.
In other words, mutations that cause selfing probabilities to fluctuate within
an epoch are not considered.
More details of the mathematical properties of selfing and cloning rates
in a coalescent context can be found in \citet{hartfield2016facsexcoal}.
